Density-dependent selection and the limits of relative fitness

### 2017-12-15 manuscript.tex
l

Thirdly, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.

@@

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather attempting to make sense of relative fitness in existing standard models of population growth mentioned above (e.g. \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. In the ``Model'' section and ``Analytical approximation of the density-dependent lottery'' part of the results section, we generalize the lottery model to allow population density to vary, creating a variable-density version of the Wright-Fisher model with overlapping generations.

In the ``$K$-selection and selection-dependent density'' section, we show that when our model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.

Furthermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.

@@

\subsection*{Analytical approximation of the density-dependent lottery}

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, note that the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

@@

Note that this behavior of $c$-selection arises directly from the role of $c$ as a trait determining relative competitive success in territorial contests, and as such this behavior is closely related to the classical appeal to reproductive excesses to justify Eq.~\eqref{eq:canonical}. By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).

\subsection*{Density-dependence and the strength of selection}

In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-selection''-type sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding. The population is in demographic equilibrium at the start of the sweep.}
\end{figure}

Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery. Selection on $c$ is clearly density-dependent, but since selection on $c$ does not affect density (section ``$K$-selection versus relative fitness''), selection on $c$ does not introduce density-dependence in Eq.~\eqref{eq:canonical}.

Turning to selection on $b$ and $d$, recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
\begin{equation}
\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
\end{equation}
The mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent, and so $d$ sweeps also follow the canonical relative fitness model exactly.

At first glance $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$. This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent. In words: the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.

Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above. The reason is that $b$ does not regulate density via the $b_i$ term in front of $f(\overline{b},N)$, which is divided by $\overline{b}$ and thus actually reflects relative lottery contests. The density regulating effect of $b$ actually appears in the fraction of contested territories $1-e^{-\overline{b}N/T}$. This reflects the fact that $b$ sweeps can only increase density by causing more unoccupied territories to receive propagules. The net effect on $f(\overline{b},N)$ is precisely zero. To see this, note that in a single-type equilibrium we have $b_i/\overline{b}=1$ and so, setting Eq.~\eqref{eq:bdensitydependence} to zero, we $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases. Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details). Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we can safely conclude that $b$ sweeps obey Eq.~\eqref{eq:canonical} to a very close approximation.

To summarize: (i) $c$-selection is density-dependent, but $c$ does not affect density (ii) $d$-selection is density-independent but $d$ affects density (iii) $b$ affects density and is subject to density-dependent selection. And yet pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation exactly (almost exactly for $b$).

@@

It is widely recognized that the relative fitness description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}. As a result, competition and the selection it induces can be concentrated at the juvenile phase, where a large reproductive excess allows for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities. This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In our lottery model, this can only occur if and only if types differ in $c$ as well as $b$ or $d$. Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}? To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded). The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).

By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent. This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

Note that while this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it any interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

>>>>

Third, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.

@@

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather than attempting to make sense of relative fitness in existing standard models of population growth (e.g. \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher relative-fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}, but unlike the Wright-Fisher model, generations can overlap. Our first task is to generalize the lottery model to create a variable-density version of the Wright-Fisher model with overlapping generations (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

Equipped with our new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}. We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density''). We then discuss selection on density-regulating traits, contrasting the results from the classical density-dependent selection literature with our model (section ``Density-regulating traits and the threat of strong selection''). We conclude [...]

@@

\subsection*{Analytical approximation of the variable-density lottery}

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

@@

The constant-$N$ behavior of $c$-selection arises from the role of $c$ as a trait determining relative competitive success in territorial contests. As such, this behavior is a result of a reproductive excesses Eq.~\eqref{eq:canonical}. By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).

\subsection*{Density-regulating traits and the threat of strong selection}

In the previous section we showed that selection and the regulation of population density can be independent even if population growth is density-regulated. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the $b$ and $d$ traits in our variable-density lottery model.

To see why this potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $K$ in the logistic model.

Starting from a monomorphic population, the frequency of a $\delta_i\rightarrow \delta_i(1-\epsilon)$ variant obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of a $b_i\rightarrow b_i(1+\epsilon)$ variant will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

Let us now turn to selection on $b$ and $d$ in our lottery model. Recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
\begin{equation}
\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
\end{equation}
It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent. Thus, $d$ sweeps follow the canonical relative fitness model exactly.

At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$. This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent. The source of this density-depedence is that the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.

Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above. The reason is that $b$ regulates density by controlling how many unoccupied territories receive propagules. Thus, greater $b$ means more propagules contesting territories, but also more territories being contested. The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases. Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details). Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).

@@

Summarizing the behaviour of the three traits in the variable-density lottery: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation. This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus, setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, sufficiently strong selection is all that is needed for relative fitness to fall apart in the classical density-dependent selection literature.

Our variable-density lottery model shows that this contrast is not simply a lack of ecological realism on the part of relative fitness models. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As noted in the introduction, this precludes the important possibility that selection is concentrated at a fragile juvenile stage as a result of a  reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

In our variable-density lottery, reproductive excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In the variable-density lottery, this occurs if and only if types differ in more than one trait. The $c$ and $d$ traits highlight the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is needed to pose a threat to Eq.~\eqref{eq:canonical}, and even then, the $b$ trait remarkably demonstrates that both may not be sufficient (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it any interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
